Fractional Types

### 2016-12-17 typespp.tex
question in email

\begin{figure}[ht]
\[\begin{array}{rrcll}
\unitepl :&  \zt \oplus \tau & \iso & \tau &: \unitipl \\
\unitepr :&  \tau \oplus \zt & \iso & \tau &: \unitipr \\
\swapp :&  \tau_1 \oplus \tau_2 & \iso & \tau_2 \oplus \tau_1 &: \swapp \\
\assoclp :&  \tau_1 \oplus (\tau_2 \oplus \tau_3) & \iso & (\tau_1 \oplus \tau_2) \oplus \tau_3
  &: \assocrp \\
\\
\unitetl :&  \ot \otimes \tau & \iso & \tau &: \unititl \\
\unitetr :&  \tau \otimes \ot & \iso & \tau &: \unititr \\
\swapt :&  \tau_1 \otimes \tau_2 & \iso & \tau_2 \otimes \tau_1 &: \swapt \\
\assoclt :&  \tau_1 \otimes (\tau_2 \otimes \tau_3) & \iso & (\tau_1 \otimes \tau_2) \otimes \tau_3
  &: \assocrt \\
\\
\absorbr :&~ \zt \otimes \tau & \iso & \zt &: \factorzl \\
\absorbl :&~ \tau \otimes \zt & \iso & \zt &: \factorzr \\
\dist :&~ (\tau_1 \oplus \tau_2) \otimes \tau_3 &
  \iso & (\tau_1 \otimes \tau_3) \oplus (\tau_2 \otimes \tau_3)~ &: \factor \\
\distl :&~ \tau_1 \otimes (\tau_2 \oplus \tau_3) &
  \iso & (\tau_1 \otimes \tau_2) \oplus (\tau_1 \otimes \tau_3)~ &: \factorl
\end{array}\]

@@

\begin{figure}[ht]
\[\begin{array}{c}
\Rule{}
{c : \tau_1 \iso \tau_2}
{\jdg{}{}{\idisotwo : c \isotwo c}}
{}
~
\Rule{}
{c_1,c_2,c_3 : \tau_1 \iso \tau_2 \quad \alpha_1 : c_1 \isotwo c_2 \quad \alpha_2 : c_2 \isotwo c_3}
{\jdg{}{}{\alpha_1~\transtwo~\alpha_2 : c_1 \isotwo c_3}}
{}
\\
\\
\Rule{}
{c_1 : \tau_1 \iso \tau_2 \quad c_2 : \tau_2 \iso \tau_3 \quad c_3 : \tau_3 \iso \tau_4}
{\jdg{}{}{\assocdl : c_1 \odot (c_2 \odot c_3) \isotwo (c_1 \odot c_2) \odot c_3 : \assocdr}}
{}
\\
\\
\Rule{}
{c : \tau_1 \iso \tau_2}
{\jdg{}{}{\idldl : \idiso \odot c \isotwo c : \idldr}}
{}
~
\Rule{}
{c : \tau_1 \iso \tau_2}
{\jdg{}{}{\idrdl : c \odot \idiso \isotwo c : \idrdr}}
{}
\\
\\
\Rule{}
{c : \tau_1 \iso \tau_2}
{\jdg{}{}{\rinvdl : ~! c \odot c \isotwo \idiso : \rinvdr}}
{}
~
\Rule{}
{c : \tau_1 \iso \tau_2}
{\jdg{}{}{\linvdl : c ~\odot ~! c \isotwo \idiso : \linvdr}}
{}
\\
\\
\Rule{}
{}
{\jdg{}{}{\sumid : \idiso\oplus\idiso \isotwo \idiso : \splitid}}
{}
\\
\\
\Rule{}
{c_1 : \tau_5\iso\tau_1 \quad c_2 : \tau_6\iso\tau_2 \quad c_3 :
    \tau_1\iso\tau_3 \quad c_4 : \tau_2\iso\tau_4}
{\jdg{}{}{\homps : (c_1\odot c_3)\oplus(c_2\odot c_4) \isotwo
    (c_1\oplus c_2) \odot (c_3\oplus c_4) : \homsp }}
{}
\\
\\
\Rule{}
{c_1,c_3 : \tau_1 \iso \tau_2 \quad c_2,c_4 : \tau_2 \iso \tau_3 \quad
  \alpha_1 : c_1 \isotwo c_3 \quad \alpha_2 : c_2 \isotwo c_4}
{\jdg{}{}{\alpha_1 ~\respstwo~ \alpha_2 : c_1 \odot c_2 \isotwo c_3 \odot c_4}}
{}
\\
\\
\Rule{}
{c_1,c_3 : \tau_1 \iso \tau_2 \quad c_2,c_4 : \tau_2 \iso \tau_3 \quad
  \alpha_1 : c_1 \isotwo c_3 \quad \alpha_2 : c_2 \isotwo c_4}
{\jdg{}{}{\respptwo ~\alpha_1 ~\alpha_2 : c_1 \oplus c_2 \isotwo c_3 \oplus c_4}}
{}
\\
\\
\Rule{}
{c_1,c_3 : \tau_1 \iso \tau_2 \quad c_2,c_4 : \tau_2 \iso \tau_3 \quad
  \alpha_1 : c_1 \isotwo c_3 \quad \alpha_2 : c_2 \isotwo c_4}
{\jdg{}{}{\respttwo ~\alpha_1 ~\alpha_2 : c_1 \otimes c_2 \isotwo c_3 \otimes c_4}}
{}
\end{array}\]
Each 2-combinator $\alpha$ has an inverse $2!~\alpha$, e.g, $2!~\assocdl=\assocdr$,
$2!(\alpha_1~\transtwo~\alpha_2) = (2!~\alpha_2)~\transtwo~(2!~\alpha_1)$, etc.
\caption{$\Pi$ 2-combinators (excerpt)~\cite{Carette2016}
\label{pi-combinators2}}
\end{figure}

\begin{figure}[ht]
{\footnotesize
\[\begin{array}{cc}
\begin{array}[t]{rlcl}
\evalone{\unitepl}{&(\inr{v})} &=& v \\
\evalone{\unitipl}{&v} &=& \inr{v} \\
\evalone{\unitepr}{&(\inl{v})} &=& v \\
\evalone{\unitipr}{&v} &=& \inl{v} \\
\evalone{\swapp}{&(\inl{v})} &=& \inr{v} \\
\evalone{\swapp}{&(\inr{v})} &=& \inl{v} \\
\evalone{\assoclp}{&(\inl{v})} &=& \inl{(\inl{v})} \\
\evalone{\assoclp}{&(\inr{(\inl{v})})} &=& \inl{(\inr{v})} \\
\evalone{\assoclp}{&(\inr{(\inr{v})})} &=& \inr{v} \\
\evalone{\assocrp}{&(\inl{(\inl{v})})} &=& \inl{v} \\
\evalone{\assocrp}{&(\inl{(\inr{v})})} &=& \inr{(\inl{v})} \\
\evalone{\assocrp}{&(\inr{v})} &=& \inr{(\inr{v})}
\end{array} &
\begin{array}[t]{rlcl}
\evalone{\unitetl}{&(\unitv,v)} &=& v \\
\evalone{\unititl}{&v} &=& (\unitv,v) \\
\evalone{\unitetr}{&(v,\unitv)} &=& v \\
\evalone{\unititr}{&v} &=& (v,\unitv) \\
\evalone{\swapt}{&(v_1,v_2)} &=& (v_2,v_1) \\
\evalone{\assoclt}{&(v_1,(v_2,v_3))} &=& ((v_1,v_2),v_3) \\
\evalone{\assocrt}{&((v_1,v_2),v_3)} &=& (v_1,(v_2,v_3))
\end{array} \\
\\
\begin{array}[t]{rlcl}
\evalone{\absorbr}{&(v,\_)} &=& v \\
\evalone{\absorbl}{&(\_,v)} &=& v \\
\evalone{\dist}{&(\inl{v_1},v_3)} &=& \inl{(v_1,v_3)} \\
\evalone{\dist}{&(\inr{v_2},v_3)} &=& \inr{(v_2,v_3)} \\
\evalone{\factor}{&\inl{(v_1,v_3)}} &=& (\inl{v_1},v_3) \\
\evalone{\factor}{&\inr{(v_2,v_3)}} &=& (\inr{v_2},v_3) \\
\evalone{\distl}{&(v_1,\inl{v_3})} &=& \inl{(v_1,v_3)} \\
\evalone{\distl}{&(v_2,\inr{v_3})} &=& \inr{(v_2,v_3)} \\
\evalone{\factorl}{&\inl{(v_1,v_3)}} &=& (v_1,\inl{v_3}) \\
\evalone{\factorl}{&\inr{(v_2,v_3)}} &=& (v_2,\inr{v_3})
\end{array} &
\begin{array}[t]{rlcl}
\evalone{\idiso}{&v} &=& v \\
\evalone{(c_1 \odot c_2)}{&v} &=&
  \evalone{c_2}{(\evalone{c_1}{v})} \\
\evalone{(c_1 \oplus c_2)}{&(\inl{v})} &=&
  \inl{(\evalone{c_1}{v})} \\
\evalone{(c_1 \oplus c_2)}{&(\inr{v})} &=&
  \inr{(\evalone{c_2}{v})} \\
\evalone{(c_1 \otimes c_2)}{&(v_1,v_2)} &=&
  (\evalone{c_1}v_1, \evalone{c_2}v_2)
\end{array}
\end{array}\]
\caption{\label{opsem}$\Pi$ operational semantics}
}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Products of Iteration and Symmetry Groupoids}

@@

\begin{figure}[ht]
\[\begin{array}{rrcll}
\unitetl/ :&  \order{\idiso} \circledast \mathbb{T} & \leftrightspoon & \mathbb{T} &: \unititl/ \\
\unitetr/ :&  \mathbb{T} \circledast \order{\idiso} & \leftrightspoon & \mathbb{T} &: \unititr/ \\
\swapt/ :&  \mathbb{T}_1 \circledast \mathbb{T}_2 & \leftrightspoon & \mathbb{T}_2 \circledast \mathbb{T}_1 &: \swapt/ \\
\assoclt/ :&  \mathbb{T}_1 \circledast (\mathbb{T}_2 \circledast \mathbb{T}_3) & \leftrightspoon & (\mathbb{T}_1 \circledast \mathbb{T}_2) \circledast \mathbb{T}_3
  &: \assocrt/ \\
\\
\eta- :& \order{\idiso} & \leftrightspoon & \iorder{c} \circledast \order{c} &: \epsilon- \\
\eta+ :& \order{\idiso} & \leftrightspoon & \order{c} \circledast \iorder{c} &: \epsilon+ \\
\end{array}\]

@@

Each combinator $\rho$ has an inverse.
\caption{$\Pi^/$ fractional combinators}
\label{pifrac:comb}
\end{figure}

We call the new extension $\Pi^/$. We begin by adding three new syntactic categories to the definition of $\Pi$ in Sec.~\ref{opsempi}:

\begin{definition}[$\Pi^/$]
\label{def:pifrac}
The syntax of $\Pi^/$ is given by the following categories:
\[\begin{array}{lrcl}
(\textrm{Types}) & \tau &::=& [\textit{see Def.~\ref{def:pi}}] \\
(\textrm{Values}) & v &::=& [\textit{see Def.~\ref{def:pi}}] \\
(\textrm{1-combinators}) & c &:& \tau_1 \iso \tau_2 ~::= [\textit{see Def.~\ref{def:pi}}] \\
(\textrm{2-combinators}) & \alpha &:& c_1 \isotwo c_2 \mbox{~where~}
  c_1, c_2 : \tau_1 \iso \tau_2 ~::= [\textit{see Def.~\ref{def:pi}}] \\
\\
(\textrm{Fractional Types}) & \mathbb{T} &::=&
  \order{c} \alt \iorder{c} \alt \mathbb{T}_1 \circledast \mathbb{T}_2 \\
 (\textrm{Fractional Values}) & \mathbb{V} &::=&
  c^k \alt 1/c^k \alt \langle \mathbb{V},\mathbb{V} \rangle \\
 (\textrm{/-combinators}) & \rho &:& \mathbb{T}_1 \leftrightspoon \mathbb{T}_2 ~::=
  [\textit{see Fig.~\ref{pifrac:comb}}]
\end{array}\]
\end{definition}

The syntactic category $\ft$ of fractional types introduces type expressions for iteration groupoids, symmetry groupoids, and their products. The combinators that relate these fractional types are in Fig.~\ref{pifrac:comb}. The bottom group includes identity and sequential composition~$\circledcirc$ combinators and ensures that the combinators can be applied anywhere inside a product $\circledast$. There is also a combinator that relates two types $\order{c_1}$ and $\order{c_2}$ via $\order{\alpha}$ whenever $\alpha : c_1 \isotwo c_2$. Indeed if $c_1$ and $c_2$ are considered equivalent via a 2-combinator then we also consider their iteration groupoids to be equivalent. The top group ensures that $\circledast$ carries a symmetric monoidal structure with combinators witnessing the unit, commutativity, and associativity properties. The last two pairs combinators $\eta-/\epsilon-$ and $\eta+/\epsilon+$ are inspired from the definition of compact closed categories~\cite{ccc} which are
symmetric monoidal categories in which every object has a dual. They are the ones that witness the ``fractional'' nature of symmetry groupoids: they are indeed the multiplicative inverses of iteration groupoids.

@@

@@

We implement an example zig-zag, which use synchr⋆
to synchronize the \emph{tangled} values created by $\eta$ and $\epsilon$.
The circuit of zig-zag is shown in figure \ref{fig:zigzag} which
shows how the input of type $\#c$ is swapped by synchr⋆.

\begin{figure}[bht]
\begin{tikzpicture}[scale=0.9,every node/.style={scale=0.9},highlight/.style={line width=4pt,orange!70}]
  \draw (0,0) -- (1,0) -- (1,2) -- (0,2) -- cycle;
  \node at (0.5,1) {$\textsf{unit}_\times$};
  \path (-1.1,1) edge node[above] {${\textsf{\#c}}$} (0,1);
  \path (1,1.75) edge node[above] {$\mathbb{1}$} (1.6,1.75);
  \path [highlight] (1,0.25) edge node[above] {} (4,0.25);
  \path (1,0.25) edge node[above] {${\textsf{\#c}}$} (4,0.25);
  \draw (1.6,1) -- (2.6,1) -- (2.6,2.8) -- (1.6,2.8) -- cycle;
  \node at (2.1,1.9) {$\eta_{\textsf{c}}$};
  \draw[dashed] (2.8,2.6) -- (2.8,1.25);
  \path [highlight] (2.6,2.6) edge node[above] {} (4,2.6);
  \path (2.6,2.6) edge node[above] {${\textsf{\#c}}$} (4,2.6);
  \path [highlight] (2.6,1.25) edge node[above] {} (4,1.25);
  \path (2.6,1.25) edge node[above] {$1/\textsf{\#c}$} (4,1.25);
  \draw (4,0) -- (5.5,0) -- (5.5,3) -- (4,3) -- cycle;
  \node at (4.75,1.5) {$synch_{\textsf{c}}$};
  \path [highlight] (5.5,1.25) edge node[above] {} (7,1.25);
  \path (5.5,1.25) edge node[above] {$1/\#c$} (7,1.25);
  \path [highlight] (5.5,2.6) edge node[above] {} (9,2.6);
  \path (5.5,2.6) edge node[above] {$\#c$} (9,2.6);
  \draw (7,0) -- (8,0) -- (8,1.6) -- (7,1.6) -- cycle;
  \node at (7.5,0.8) {$\epsilon_c$};
  \draw[dashed] (6.8,0.2) -- (6.8,1.25);
  \path (8,1) edge node[above] {$\mathbb{1}$} (9,1);
  \path [highlight] (5.5,0.2) edge node[above] {} (7,0.2);
  \path (5.5,0.2) edge node[above] {$\#c$} (7,0.2);
  \draw (9,0.8) -- (10,0.8) -- (10,2.8) -- (9,2.8) -- cycle;
  \node at (9.5,1.8) {$\textsf{unit}_\times$};
  \path (10,1.8) edge node [above] {$\#c$} (11,1.8);
\end{tikzpicture}
\caption{\label{fig:zigzag} Zig-zag}
\end{figure}

>>>>

\begin{figure}[t]
\[\begin{array}{rrcll}
\unitepl :&  \zt \oplus \tau & \iso & \tau &: \unitipl \\
\unitepr :&  \tau \oplus \zt & \iso & \tau &: \unitipr \\
\swapp :&  \tau_1 \oplus \tau_2 & \iso & \tau_2 \oplus \tau_1 &: \swapp \\
\assoclp :&  \tau_1 \oplus (\tau_2 \oplus \tau_3) & \iso & (\tau_1 \oplus \tau_2) \oplus \tau_3
  &: \assocrp \\
\\
\unitetl :&  \ot \otimes \tau & \iso & \tau &: \unititl \\
\unitetr :&  \tau \otimes \ot & \iso & \tau &: \unititr \\
\swapt :&  \tau_1 \otimes \tau_2 & \iso & \tau_2 \otimes \tau_1 &: \swapt \\
\assoclt :&  \tau_1 \otimes (\tau_2 \otimes \tau_3) & \iso & (\tau_1 \otimes \tau_2) \otimes \tau_3
  &: \assocrt \\
\\
\absorbr :&~ \zt \otimes \tau & \iso & \zt &: \factorzl \\
\absorbl :&~ \tau \otimes \zt & \iso & \zt &: \factorzr \\
\dist :&~ (\tau_1 \oplus \tau_2) \otimes \tau_3 &
  \iso & (\tau_1 \otimes \tau_3) \oplus (\tau_2 \otimes \tau_3)~ &: \factor \\
\distl :&~ \tau_1 \otimes (\tau_2 \oplus \tau_3) &
  \iso & (\tau_1 \otimes \tau_2) \oplus (\tau_1 \otimes \tau_3)~ &: \factorl
\end{array}\]

@@

\begin{figure}[t]
\[\begin{array}{c}
\Rule{}
{c : \tau_1 \iso \tau_2}
{\jdg{}{}{\idisotwo : c \isotwo c}}
{}
~
\Rule{}
{c_1,c_2,c_3 : \tau_1 \iso \tau_2 \quad \alpha_1 : c_1 \isotwo c_2 \quad \alpha_2 : c_2 \isotwo c_3}
{\jdg{}{}{\alpha_1~\transtwo~\alpha_2 : c_1 \isotwo c_3}}
{}
\\
\\
\Rule{}
{c_1 : \tau_1 \iso \tau_2 \quad c_2 : \tau_2 \iso \tau_3 \quad c_3 : \tau_3 \iso \tau_4}
{\jdg{}{}{\assocdl : c_1 \odot (c_2 \odot c_3) \isotwo (c_1 \odot c_2) \odot c_3 : \assocdr}}
{}
\\
\\
\Rule{}
{c : \tau_1 \iso \tau_2}
{\jdg{}{}{\idldl : \idiso \odot c \isotwo c : \idldr}}
{}
~
\Rule{}
{c : \tau_1 \iso \tau_2}
{\jdg{}{}{\idrdl : c \odot \idiso \isotwo c : \idrdr}}
{}
\\
\\
\Rule{}
{c : \tau_1 \iso \tau_2}
{\jdg{}{}{\rinvdl : ~! c \odot c \isotwo \idiso : \rinvdr}}
{}
~
\Rule{}
{c : \tau_1 \iso \tau_2}
{\jdg{}{}{\linvdl : c ~\odot ~! c \isotwo \idiso : \linvdr}}
{}
\\
\\
\Rule{}
{}
{\jdg{}{}{\sumid : \idiso\oplus\idiso \isotwo \idiso : \splitid}}
{}
\\
\\
\Rule{}
{c_1 : \tau_5\iso\tau_1 \quad c_2 : \tau_6\iso\tau_2 \quad c_3 :
    \tau_1\iso\tau_3 \quad c_4 : \tau_2\iso\tau_4}
{\jdg{}{}{\homps : (c_1\odot c_3)\oplus(c_2\odot c_4) \isotwo
    (c_1\oplus c_2) \odot (c_3\oplus c_4) : \homsp }}
{}
\\
\\
\Rule{}
{c_1,c_3 : \tau_1 \iso \tau_2 \quad c_2,c_4 : \tau_2 \iso \tau_3 \quad
  \alpha_1 : c_1 \isotwo c_3 \quad \alpha_2 : c_2 \isotwo c_4}
{\jdg{}{}{\alpha_1 ~\respstwo~ \alpha_2 : c_1 \odot c_2 \isotwo c_3 \odot c_4}}
{}
\\
\\
\Rule{}
{c_1,c_3 : \tau_1 \iso \tau_2 \quad c_2,c_4 : \tau_2 \iso \tau_3 \quad
  \alpha_1 : c_1 \isotwo c_3 \quad \alpha_2 : c_2 \isotwo c_4}
{\jdg{}{}{\respptwo ~\alpha_1 ~\alpha_2 : c_1 \oplus c_2 \isotwo c_3 \oplus c_4}}
{}
\\
\\
\Rule{}
{c_1,c_3 : \tau_1 \iso \tau_2 \quad c_2,c_4 : \tau_2 \iso \tau_3 \quad
  \alpha_1 : c_1 \isotwo c_3 \quad \alpha_2 : c_2 \isotwo c_4}
{\jdg{}{}{\respttwo ~\alpha_1 ~\alpha_2 : c_1 \otimes c_2 \isotwo c_3 \otimes c_4}}
{}
\end{array}\]
Each 2-combinator $\alpha$ has an inverse $2!~\alpha$, e.g, $2!~\assocdl=\assocdr$,
$2!(\alpha_1~\transtwo~\alpha_2) = (2!~\alpha_2)~\transtwo~(2!~\alpha_1)$, etc.
\caption{$\Pi$ 2-combinators (excerpt)~\cite{Carette2016}
\label{pi-combinators2}}
\end{figure}

\begin{figure}[t]
{\footnotesize
\[\begin{array}{cc}
\begin{array}[t]{rlcl}
\evalone{\unitepl}{&(\inr{v})} &=& v \\
\evalone{\unitipl}{&v} &=& \inr{v} \\
\evalone{\unitepr}{&(\inl{v})} &=& v \\
\evalone{\unitipr}{&v} &=& \inl{v} \\
\evalone{\swapp}{&(\inl{v})} &=& \inr{v} \\
\evalone{\swapp}{&(\inr{v})} &=& \inl{v} \\
\evalone{\assoclp}{&(\inl{v})} &=& \inl{(\inl{v})} \\
\evalone{\assoclp}{&(\inr{(\inl{v})})} &=& \inl{(\inr{v})} \\
\evalone{\assoclp}{&(\inr{(\inr{v})})} &=& \inr{v} \\
\evalone{\assocrp}{&(\inl{(\inl{v})})} &=& \inl{v} \\
\evalone{\assocrp}{&(\inl{(\inr{v})})} &=& \inr{(\inl{v})} \\
\evalone{\assocrp}{&(\inr{v})} &=& \inr{(\inr{v})}
\end{array} &
\begin{array}[t]{rlcl}
\evalone{\unitetl}{&(\unitv,v)} &=& v \\
\evalone{\unititl}{&v} &=& (\unitv,v) \\
\evalone{\unitetr}{&(v,\unitv)} &=& v \\
\evalone{\unititr}{&v} &=& (v,\unitv) \\
\evalone{\swapt}{&(v_1,v_2)} &=& (v_2,v_1) \\
\evalone{\assoclt}{&(v_1,(v_2,v_3))} &=& ((v_1,v_2),v_3) \\
\evalone{\assocrt}{&((v_1,v_2),v_3)} &=& (v_1,(v_2,v_3))
\end{array} \\
\\
\begin{array}[t]{rlcl}
\evalone{\absorbr}{&(v,\_)} &=& v \\
\evalone{\absorbl}{&(\_,v)} &=& v \\
\evalone{\dist}{&(\inl{v_1},v_3)} &=& \inl{(v_1,v_3)} \\
\evalone{\dist}{&(\inr{v_2},v_3)} &=& \inr{(v_2,v_3)} \\
\evalone{\factor}{&\inl{(v_1,v_3)}} &=& (\inl{v_1},v_3) \\
\evalone{\factor}{&\inr{(v_2,v_3)}} &=& (\inr{v_2},v_3) \\
\evalone{\distl}{&(v_1,\inl{v_3})} &=& \inl{(v_1,v_3)} \\
\evalone{\distl}{&(v_2,\inr{v_3})} &=& \inr{(v_2,v_3)} \\
\evalone{\factorl}{&\inl{(v_1,v_3)}} &=& (v_1,\inl{v_3}) \\
\evalone{\factorl}{&\inr{(v_2,v_3)}} &=& (v_2,\inr{v_3})
\end{array} &
\begin{array}[t]{rlcl}
\evalone{\idiso}{&v} &=& v \\
\evalone{(c_1 \odot c_2)}{&v} &=&
  \evalone{c_2}{(\evalone{c_1}{v})} \\
\evalone{(c_1 \oplus c_2)}{&(\inl{v})} &=&
  \inl{(\evalone{c_1}{v})} \\
\evalone{(c_1 \oplus c_2)}{&(\inr{v})} &=&
  \inr{(\evalone{c_2}{v})} \\
\evalone{(c_1 \otimes c_2)}{&(v_1,v_2)} &=&
  (\evalone{c_1}v_1, \evalone{c_2}v_2)
\end{array}
\end{array}\]
\caption{\label{opsem}$\Pi$ operational semantics}
}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Products of Iteration and Symmetry Groupoids}
\label{sub:products}

@@

\begin{figure}[t]
\[\begin{array}{rrcll}
\unitetl/ :&  \order{\idiso} \circledast \mathbb{T} & \leftrightspoon & \mathbb{T} &: \unititl/ \\
\unitetr/ :&  \mathbb{T} \circledast \order{\idiso} & \leftrightspoon & \mathbb{T} &: \unititr/ \\
\swapt/ :&  \mathbb{T}_1 \circledast \mathbb{T}_2 & \leftrightspoon & \mathbb{T}_2 \circledast \mathbb{T}_1 &: \swapt/ \\
\assoclt/ :&  \mathbb{T}_1 \circledast (\mathbb{T}_2 \circledast \mathbb{T}_3) & \leftrightspoon & (\mathbb{T}_1 \circledast \mathbb{T}_2) \circledast \mathbb{T}_3
  &: \assocrt/ \\
\\
\eta- :& \order{\idiso} & \leftrightspoon & \iorder{c} \circledast \order{c} &: \epsilon- \\
\eta+ :& \order{\idiso} & \leftrightspoon & \order{c} \circledast \iorder{c} &: \epsilon+ \\
\end{array}\]

@@

Each combinator $\rho$ has an inverse.
\caption{$\Pi^/$ fractional combinators (non-dependent version)}
\label{pifrac:comb}
\end{figure}

%%%%%
\subsection{Non-Dependent Version}

As a first approximation to the language we seek, we consider adding three new
syntactic categories to the definition of $\Pi$ in Sec.~\ref{opsempi}:

\begin{definition}[Non-dependent $\Pi^/$]
\label{def:pifrac}
The syntax of (non-dependent) $\Pi^/$ is given by the following categories:
\[\begin{array}{lrcl}
(\textrm{Types}) & \tau &::=& [\textit{see Def.~\ref{def:pi}}] \\
(\textrm{Values}) & v &::=& [\textit{see Def.~\ref{def:pi}}] \\
(\textrm{1-combinators}) & c &:& \tau_1 \iso \tau_2 ~::= [\textit{see Def.~\ref{def:pi}}] \\
(\textrm{2-combinators}) & \alpha &:& c_1 \isotwo c_2 \mbox{~where~}
  c_1, c_2 : \tau_1 \iso \tau_2 ~::= [\textit{see Def.~\ref{def:pi}}] \\
\\
(\textrm{Fractional Types}) & \mathbb{T} &::=&
  \order{c} \alt \iorder{c} \alt \mathbb{T}_1 \circledast \mathbb{T}_2 \\
 (\textrm{Fractional Values}) & \mathbb{V} &::=&
  c^k \alt 1/c^k \alt \langle \mathbb{V},\mathbb{V} \rangle \\
 (\textrm{/-combinators}) & \rho &:& \mathbb{T}_1 \leftrightspoon \mathbb{T}_2 ~::=
  [\textit{see Fig.~\ref{pifrac:comb}}]
\end{array}\]
\end{definition}

The syntactic category $\ft$ of fractional types introduces type expressions for
iteration groupoids, symmetry groupoids, and their products. The combinators
that relate these fractional types are in Fig.~\ref{pifrac:comb}. The bottom
group includes identity and sequential composition~$\circledcirc$ combinators
and ensures that the combinators can be applied anywhere inside a product
$\circledast$. There is also a combinator that relates two types $\order{c_1}$
and $\order{c_2}$ via $\order{\alpha}$ whenever $\alpha : c_1 \isotwo
c_2$.
Indeed if $c_1$ and $c_2$ are considered equivalent via a 2-combinator then we
also consider their iteration groupoids to be equivalent. The top group ensures
that $\circledast$ carries a symmetric monoidal structure with combinators
witnessing the unit, commutativity, and associativity properties. The last two
pairs combinators $\eta-/\epsilon-$ and $\eta+/\epsilon+$ are inspired from the
definition of compact closed categories~\cite{ccc} which are symmetric monoidal
categories in which every object has a dual. They are the ones that witness the
``fractional'' nature of symmetry groupoids as motivated in Sec.~\ref{sub:products}.

@@

%%%%%
\subsection{Pointed Types}

To understand the challenge in designing an operational semantics for $\eta$ and
$\epsilon$ in the non-dependent setting, we consider the ``zig-zag'' circuit in
Fig.~\ref{fig:zigzag}. Let the combinator $c$ to be of order greater than or
equal to 2, i.e., the type $\order{c}$ has at least two distinct values $c^i$
and $c^j$ and the type $\iorder{c}$ has at least two distinct values $1/c^i$ and
$1/c^j$. The use of $\eta+$ has then at least four choices; as discussed at the
end of Sec.~\ref{sub:products}, $\eta+$ will internalize some arbitrary but
fixed choice and might for example generate the pair
$\langle c^i , 1/c^i \rangle$. No matter what choice is internalized in $\eta+$,
there is at least one different incoming value $c^j$ of type $\order{c}$. In
that case, the combinator $\epsilon-$ encounters $\langle 1/c^i , c^j
\rangle$. If our operational semantics allows $\epsilon-$ to annihilate
arbitrary pairs, the net effect of the circuit would be to map $c^j$ to $c^i$
for arbitrary $i$ and $j$.

\begin{figure}[bht]
\begin{tikzpicture}[scale=0.9,every node/.style={scale=0.7},highlight/.style={line width=4pt,orange!70}]
  \draw (0,0) -- (1,0) -- (1,2) -- (0,2) -- cycle;
  \node at (0.5,1) {\footnotesize{$\unitetl/$}};
  \path (-1.1,1) edge node[above] {\footnotesize{${\textsf{\#c}}$}} (0,1);
  \path (1,1.75) edge node[above] {\footnotesize{$\order{\idiso}$}} (1.6,1.75);
  \path [highlight] (1,0.25) edge node[above] {} (4,0.25);
  \path (1,0.25) edge node[above] {\footnotesize{${\textsf{\#c}}$}} (4,0.25);
  \draw (1.6,1) -- (2.6,1) -- (2.6,2.8) -- (1.6,2.8) -- cycle;
  \node at (2.1,1.9) {\footnotesize{$\eta+$}};
  \draw[dashed] (2.8,2.6) -- (2.8,1.25);
  \path [highlight] (2.6,2.6) edge node[above] {} (4,2.6);
  \path (2.6,2.6) edge node[above] {\footnotesize{${\textsf{\#c}}$}} (4,2.6);
  \path [highlight] (2.6,1.25) edge node[above] {} (4,1.25);
  \path (2.6,1.25) edge node[above] {\footnotesize{$1/\textsf{\#c}$}} (4,1.25);
  \draw (4,0) -- (5.5,0) -- (5.5,3) -- (4,3) -- cycle;
  \node at (4.75,1.5) {$\footnotesize{\assocrt/}$};
  \path [highlight] (5.5,1.25) edge node[above] {} (7,1.25);
  \path (5.5,1.25) edge node[above] {$\footnotesize{1/\#c}$} (7,1.25);
  \path [highlight] (5.5,2.6) edge node[above] {} (9,2.6);
  \path (5.5,2.6) edge node[above] {\footnotesize{$\#c$}} (9,2.6);
  \draw (7,0) -- (8,0) -- (8,1.6) -- (7,1.6) -- cycle;
  \node at (7.5,0.8) {\footnotesize{$\epsilon-$}};
  \draw[dashed] (6.8,0.2) -- (6.8,1.25);
  \path (8,1) edge node[above] {\footnotesize{$\order{\idiso}$}} (9,1);
  \path [highlight] (5.5,0.2) edge node[above] {} (7,0.2);
  \path (5.5,0.2) edge node[above] {\footnotesize{$\#c$}} (7,0.2);
  \draw (9,0.8) -- (10,0.8) -- (10,2.8) -- (9,2.8) -- cycle;
  \node at (9.5,1.8) {\footnotesize{$\unitetr/$}};
  \path (10,1.8) edge node [above] {\footnotesize{$\#c$}} (11,1.8);
\end{tikzpicture}
\caption{\label{fig:zigzag} Zig-zag}
\end{figure}

@@
